---
title: "Comparison with SAS and Stata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with SAS and Stata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(icsp2)
library(survival)
```

## Simulation Framework

We generate interval-censored data from a proportional hazards model with
Weibull baseline, allowing per-stratum baseline parameters and treatment
effects. We then compare:
 
- **Log-rank test**: `ic_logrank()` vs SAS PROC ICLIFETEST
- **PH model fit**: `ic_sp_ph()` vs Stata `stintcox`

### Simulation wrapper

```{r sim-wrapper}
#' Simulate interval-censored trial data with optional stratification
#'
#' @param n_per_arm Sample size per treatment arm (total N = 2 * n_per_arm * n_strata)
#' @param b_trt Treatment effect (log-HR). Scalar or vector of length n_strata.
#' @param b_age Coefficient for continuous age covariate. NULL to omit age.
#' @param strata_params List of lists, one per stratum, each with `shape` and `scale`.
#'   If NULL, no stratification (single stratum with Weibull(2,2)).
#' @param inspections Number of inspection times
#' @param inspectLength Maximum gap between inspections
#' @param seed Random seed
#' @return A data.frame with columns: l, u, trt, and optionally age, strata
sim_ic_trial <- function(
  n_per_arm = 200,
  b_trt = 0.5,
  b_age = NULL,
  strata_params = NULL,
  inspections = 4,
  inspectLength = 1.5,
  seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  # Default: single stratum with Weibull(2, 2)
  if (is.null(strata_params)) {
    strata_params <- list(list(shape = 2, scale = 2))
  }

  n_strata <- length(strata_params)
  b_trt <- rep_len(b_trt, n_strata)

  all_data <- vector("list", n_strata)

  for (s in seq_len(n_strata)) {
    n_s <- 2 * n_per_arm
    trt <- rep(c(0, 1), each = n_per_arm)
    lin_pred <- trt * b_trt[s]

    # Optional continuous covariate
    age <- NULL
    if (!is.null(b_age)) {
      age <- rnorm(n_s)
      lin_pred <- lin_pred + age * b_age
    }

    # Generate exact event times from PH Weibull
    shape_s <- strata_params[[s]]$shape
    scale_s <- strata_params[[s]]$scale
    raw_p <- runif(n_s)
    nu <- exp(lin_pred)
    adj_p <- 1 - raw_p^(1 / nu)
    true_times <- qweibull(adj_p, shape = shape_s, scale = scale_s)

    # Interval censoring mechanism
    obs_times <- runif(n_s, min = 0, max = inspectLength)
    l <- rep(0, n_s)
    u <- rep(Inf, n_s)

    caught <- true_times < obs_times
    u[caught] <- obs_times[caught]
    l[!caught] <- obs_times[!caught]

    if (inspections > 1) {
      for (k in 2:inspections) {
        old_obs <- obs_times
        obs_times <- old_obs + runif(n_s, min = 0, max = inspectLength)
        newly_caught <- true_times >= old_obs & true_times < obs_times
        still_uncaught <- true_times >= obs_times
        u[newly_caught] <- obs_times[newly_caught]
        l[still_uncaught] <- obs_times[still_uncaught]
      }
    }

    df_s <- data.frame(l = l, u = u, trt = trt)
    if (!is.null(age)) df_s$age <- age
    if (n_strata > 1) df_s$strata <- paste0("S", s)

    all_data[[s]] <- df_s
  }

  out <- do.call(rbind, all_data)
  if ("strata" %in% names(out)) out$strata <- factor(out$strata)
  out$trt <- factor(out$trt)
  rownames(out) <- NULL
  out
}
```

## Scenario Definitions

```{r scenarios}
scenarios <- list(
  S1_null = list(
    label = "Null effect",
    b_trt = 0,
    b_age = NULL,
    strata_params = NULL,
    inspections = 4,
    inspectLength = 1.5
  ),
  S2_moderate = list(
    label = "Moderate effect (HR~1.65)",
    b_trt = 0.5,
    b_age = NULL,
    strata_params = NULL,
    inspections = 4,
    inspectLength = 1.5
  ),
  S3_strong = list(
    label = "Strong effect (HR~2.72)",
    b_trt = 1.0,
    b_age = NULL,
    strata_params = NULL,
    inspections = 5,
    inspectLength = 1.0
  ),
  S4_weak = list(
    label = "Weak effect (HR~1.22)",
    b_trt = 0.2,
    b_age = NULL,
    strata_params = list(list(shape = 1.5, scale = 3)),
    inspections = 3,
    inspectLength = 2.5
  ),
  S5_moderate_age = list(
    label = "Moderate + continuous covariate",
    b_trt = 0.5,
    b_age = 0.3,
    strata_params = NULL,
    inspections = 4,
    inspectLength = 1.5
  ),
  S6_strat_homo = list(
    label = "Stratified, same baseline, homogeneous effect",
    b_trt = 0.5,
    b_age = NULL,
    strata_params = list(
      list(shape = 2, scale = 2),
      list(shape = 2, scale = 2)
    ),
    inspections = 4,
    inspectLength = 2.0
  ),
  S7_strat_diffbase = list(
    label = "Stratified, different baselines, homogeneous effect",
    b_trt = 0.5,
    b_age = NULL,
    strata_params = list(
      list(shape = 2, scale = 2),
      list(shape = 1, scale = 4)
    ),
    inspections = 3,
    inspectLength = 2.0
  ),
  S8_strat_hetero = list(
    label = "Stratified, different baselines, heterogeneous effect",
    b_trt = c(0.7, 0),
    b_age = NULL,
    strata_params = list(
      list(shape = 2, scale = 1.5),
      list(shape = 3, scale = 3)
    ),
    inspections = 4,
    inspectLength = 1.0
  )
)
```

## Generate Datasets

We generate 100 replications per scenario with `n_per_arm = 200`.

```{r generate-data}
n_reps <- 100
n_per_arm <- 200

set.seed(2026)
seeds <- sample.int(1e7, size = n_reps * length(scenarios))
seed_matrix <- matrix(seeds, nrow = n_reps, ncol = length(scenarios))

sim_datasets <- vector("list", length(scenarios))
names(sim_datasets) <- names(scenarios)

for (j in seq_along(scenarios)) {
  sc <- scenarios[[j]]
  sim_datasets[[j]] <- lapply(seq_len(n_reps), function(i) {
    sim_ic_trial(
      n_per_arm = n_per_arm,
      b_trt = sc$b_trt,
      b_age = sc$b_age,
      strata_params = sc$strata_params,
      inspections = sc$inspections,
      inspectLength = sc$inspectLength,
      seed = seed_matrix[i, j]
    )
  })
}
```

## Run icsp2 Analyses

### Log-rank tests

```{r logrank-analysis}
logrank_results <- vector("list", length(scenarios))
names(logrank_results) <- names(scenarios)

for (j in seq_along(scenarios)) {
  sc <- scenarios[[j]]
  has_strata <- !is.null(sc$strata_params) && length(sc$strata_params) > 1

  logrank_results[[j]] <- lapply(sim_datasets[[j]], function(dat) {
    if (has_strata) {
      fml <- Surv(l, u, type = "interval2") ~ trt + strata(strata)
    } else {
      fml <- Surv(l, u, type = "interval2") ~ trt
    }
    ic_logrank(fml, data = dat, n_samples = 10000)
  })
}

# Extract p-values
logrank_pvalues <- lapply(logrank_results, function(res_list) {
  vapply(res_list, function(x) x$p.value, numeric(1))
})

logrank_statistics <- lapply(logrank_results, function(res_list) {
  vapply(res_list, function(x) x$statistic, numeric(1))
})

```


```{r}
# read sas files

```

### PH model fits

```{r ph-analysis}
ph_results <- vector("list", length(scenarios))
names(ph_results) <- names(scenarios)

for (j in seq_along(scenarios)) {
  sc <- scenarios[[j]]
  has_strata <- !is.null(sc$strata_params) && length(sc$strata_params) > 1
  has_age <- !is.null(sc$b_age)

  ph_results[[j]] <- lapply(sim_datasets[[j]], function(dat) {
    rhs <- "trt"
    if (has_age) rhs <- paste(rhs, "+ age")
    if (has_strata) rhs <- paste(rhs, "+ strata(strata)")
    fml <- as.formula(paste("Surv(l, u, type = 'interval2') ~", rhs))
    ic_sp_ph(fml, data = dat)
  })
}
})
# Extract coefficients
ph_coefs <- lapply(ph_results, function(res_list) {
  t(vapply(
    res_list,
    function(x) x$coefficients,
    numeric(length(res_list[[1]]$coefficients))
  ))
})
```

## Baseline Survival Plots

For one representative dataset per scenario, we plot the estimated baseline
survival from `ic_sp_ph()` against the true Weibull survival. For stratified
scenarios, each stratum is plotted separately.

```{r baseline-plots, fig.width=8, fig.height=10}
par(mfrow = c(4, 2), mar = c(4, 4, 2, 1))

for (j in seq_along(scenarios)) {
  sc <- scenarios[[j]]
  fit <- ph_results[[j]][[1]]  # first replication

  strata_pars <- sc$strata_params
  if (is.null(strata_pars)) strata_pars <- list(list(shape = 2, scale = 2))
  n_strata <- length(strata_pars)

  # Determine x-axis range from the fit
  x_max <- max(unlist(lapply(fit$intervals, function(int) max(int[is.finite(int)]))))

  plot(fit,
    main = paste0("S", j, ": ", sc$label),
    # xlim = c(0, x_max),
    col = rep(seq_len(n_strata), each = 2),
    lwd = 2
  )

  # Overlay true baseline survival for each stratum
  t_seq <- seq(0, x_max, length.out = 200)
  for (s in seq_len(n_strata)) {
    true_surv <- pweibull(
      t_seq,
      shape = strata_pars[[s]]$shape,
      scale = strata_pars[[s]]$scale,
      lower.tail = FALSE
    )
    lines(t_seq, true_surv, col = s, lty = 2, lwd = 2)
  }
  legend("topright",
    legend = c("Estimated", "True"),
    lty = c(1, 2), lwd = 2, cex = 0.7, bty = "n"
  )
}
```

## Export Data for SAS and Stata

Export each dataset as a CSV for analysis in external software.

```{r export-data}
export_dir <- "./local/sim_data"
dir.create(export_dir, recursive = TRUE, showWarnings = FALSE)

for (j in seq_along(scenarios)) {
  sc_name <- names(scenarios)[j]
  sc_dir <- file.path(export_dir, sc_name)
  dir.create(sc_dir, recursive = TRUE, showWarnings = FALSE)
  for (i in seq_along(sim_datasets[[j]])) {
    write.csv(
      sim_datasets[[j]][[i]],
      file = file.path(sc_dir, sprintf("rep_%03d.csv", i)),
      row.names = FALSE
    )
  }
}
```

## SAS Code Template (PROC ICLIFETEST)

The following SAS code can be used to process the exported datasets:

```
/* SAS macro for running PROC ICLIFETEST on all replications */
%macro run_iclifetest(scenario, n_reps=100, has_strata=0);
  %do i = 1 %to &n_reps;
    proc import datafile="sim_data/&scenario/rep_%sysfunc(putn(&i, z3.)).csv"
      out=work.rep_&i dbms=csv replace; run;

    /* Convert u from character to numeric (Inf -> .) */
    data work.rep_&i;
      set work.rep_&i(rename=(u=u_char));
      if upcase(u_char) = "INF" then u = .;
      else u = input(u_char, best.);
      drop u_char;
    run;

    proc iclifetest data=work.rep_&i
      %if &i = 1 %then %do;
        plots=(survival) outsurv=work.surv_1
      %end;
      ;
      time (l, u);
      %if &has_strata = 1 %then %do;
        strata strata;
      %end;
      test trt;
      ods output LogRankTest=work.lr_&i;
    run;

    /* Add replication index */
    data work.lr_&i;
      set work.lr_&i;
      rep = &i;
    run;
  %end;

  /* Stack all log-rank results and export */
  data work.&scenario._lr;
    set work.lr_1 - work.lr_&n_reps;
  run;

  proc export data=work.&scenario._lr
    outfile="sim_data/&scenario/sas_logrank.csv"
    dbms=csv replace;
  run;

  /* Export NPMLE survival from first replication */
  proc export data=work.surv_1
    outfile="sim_data/&scenario/sas_npmle_surv.csv"
    dbms=csv replace;
  run;
%mend;

%run_iclifetest(S1_null, has_strata=0);
%run_iclifetest(S2_moderate, has_strata=0);
%run_iclifetest(S3_strong, has_strata=0);
%run_iclifetest(S4_weak, has_strata=0);
%run_iclifetest(S5_moderate_age, has_strata=0);
%run_iclifetest(S6_strat_homo, has_strata=1);
%run_iclifetest(S7_strat_diffbase, has_strata=1);
%run_iclifetest(S8_strat_hetero, has_strata=1);
```

## Stata Code Template (stintcox)

```
/* Stata do-file for running stintcox on all replications */
program define run_stintcox
  args scenario n_reps has_strata has_age

  local n_coefs = 1 + ("`has_age'" == "1")

  forvalues i = 1/`n_reps' {
    local fname "sim_data/`scenario'/rep_`=string(`i', "%03.0f")'.csv"
    import delimited "`fname'", clear

    * Handle right-censored (Inf) values
    replace u = . if u == "Inf"
    destring u, replace

    stintcox trt `=cond("`has_age'"=="1", "age", "")' ///
      , interval(l u) `=cond("`has_strata'"=="1", "strata(strata)", "")'

    * Store coefficients and SEs
    matrix b = e(b)
    matrix V = e(V)
    matrix se = J(1, `n_coefs', .)
    forvalues c = 1/`n_coefs' {
      matrix se[1, `c'] = sqrt(V[`c', `c'])
    }

    if `i' == 1 {
      matrix all_b = b
      matrix all_se = se
    }
    else {
      matrix all_b = all_b \ b
      matrix all_se = all_se \ se
    }

    * Save baseline survival from first replication
    if `i' == 1 {
      predict double surv0, basesurv
      predict double haz0, basehaz
      gen double time_grid = _t
      keep time_grid surv0 haz0
      duplicates drop
      sort time_grid
      export delimited "sim_data/`scenario'/stata_baseline.csv", replace
    }
  }

  * Export coefficient and SE matrices to CSV
  clear
  svmat all_b, names(b)
  svmat all_se, names(se)
  gen rep = _n
  export delimited "sim_data/`scenario'/stata_coefs.csv", replace
end

run_stintcox S1_null 100 0 0
run_stintcox S2_moderate 100 0 0
run_stintcox S3_strong 100 0 0
run_stintcox S4_weak 100 0 0
run_stintcox S5_moderate_age 100 0 1
run_stintcox S6_strat_homo 100 1 0
run_stintcox S7_strat_diffbase 100 1 0
run_stintcox S8_strat_hetero 100 1 0
```

## Summary of Results

### Load external results

```{r load-external}
sim_data_dir <- "./comparison_sim/"

# --- SAS log-rank results ---
sas_logrank <- lapply(names(scenarios), function(sc_name) {
  f <- file.path(sim_data_dir, "output", sc_name, "sas_logrank.csv")
  if (file.exists(f)) {
    d <- read.csv(f)
    d$scenario <- sc_name
    d
  }
})
sas_logrank <- do.call(rbind, sas_logrank)

# --- Stata coefficient results ---
stata_coefs <- lapply(names(scenarios), function(sc_name) {
  f <- file.path(sim_data_dir, "output_stata", sc_name, "stata_coefs10k.csv")
  list.files()
  if (file.exists(f)) {
    d <- read.csv(f)
    d$scenario <- sc_name
    d
  }
})
stata_coefs <- do.call(rbind, lapply(stata_coefs, \(x) x[, c("b1", "se1", "rep", "scenario")]))

# --- SAS NPMLE survival (rep 1) ---
sas_npmle <- lapply(names(scenarios), function(sc_name) {
  f <- file.path(sim_data_dir, sc_name, "sas_npmle_surv.csv")
  if (file.exists(f)) {
    d <- read.csv(f)
    d$scenario <- sc_name
    d
  }
})
sas_npmle <- do.call(rbind, sas_npmle)

# --- Stata baseline survival (rep 1) ---
stata_baseline <- lapply(names(scenarios), function(sc_name) {
  f <- file.path(sim_data_dir, sc_name, "stata_baseline.csv")
  if (file.exists(f)) {
    d <- read.csv(f)
    d$scenario <- sc_name
    d
  }
})
stata_baseline <- do.call(rbind, stata_baseline)
```

```{r}

library(ggplot2)
ggplot(
  data.frame(
    scenario = gl(8, 100, labels = names(scenarios)),
    SAS_pval = pchisq(sas_logrank$ChiSquare, df = 1,  lower.tail = FALSE),
    icsp2_pval = icsp2_logrank$icsp2_pval
  ),
  aes(
    x = (SAS_pval + icsp2_pval)/2,
    y = SAS_pval - icsp2_pval,
  ) 
) + 
  annotate("ribbon", xmin = -Inf, x=0:1, ymin = -0.001, ymax = 0.001, fill = rgb(1, 0, 0, 0.5)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~scenario, nrow = 2) +
  theme_bw() +
  labs(
    x = "(SAS + R)/2",
    y = "SAS - R"
  )
ggsave("icsp2_v_sas.png", dpi = 300, width = 9.15, height = 4.32, units = "in")
  
```


### Log-rank comparison (icsp2 vs SAS)

```{r logrank-comparison}
# Extract icsp2 statistics and p-values per replication
icsp2_logrank <- lapply(seq_along(scenarios), function(j) {
  data.frame(
    scenario = names(scenarios)[j],
    rep = seq_along(logrank_results[[j]]),
    icsp2_stat = vapply(logrank_results[[j]], function(x) x$statistic, numeric(1)),
    icsp2_pval = vapply(logrank_results[[j]], function(x) x$p.value, numeric(1))
  )
})
icsp2_logrank <- do.call(rbind, icsp2_logrank)

# Merge with SAS results if available
if (!is.null(sas_logrank) && nrow(sas_logrank) > 0) {
  lr_comparison <- merge(
    icsp2_logrank,
    sas_logrank[c("scenario", "rep", "ChiSq", "ProbChiSq")],
    by = c("scenario", "rep"),
    all.x = TRUE
  )
  names(lr_comparison)[names(lr_comparison) == "ChiSq"] <- "sas_stat"
  names(lr_comparison)[names(lr_comparison) == "ProbChiSq"] <- "sas_pval"

  # Per-scenario agreement summary
  lr_agreement <- do.call(rbind, lapply(split(lr_comparison, lr_comparison$scenario), function(d) {
    data.frame(
      scenario = d$scenario[1],
      max_abs_diff_stat = max(abs(d$icsp2_stat - d$sas_stat), na.rm = TRUE),
      max_abs_diff_pval = max(abs(d$icsp2_pval - d$sas_pval), na.rm = TRUE),
      cor_stat = cor(d$icsp2_stat, d$sas_stat, use = "complete.obs"),
      cor_pval = cor(d$icsp2_pval, d$sas_pval, use = "complete.obs")
    )
  }))
  rownames(lr_agreement) <- NULL
  lr_agreement
} else {
  icsp2_logrank
}
```

### PH coefficient comparison (icsp2 vs Stata)

```{r ph-comparison}
# icsp2 PH summary

icsp2_coef <- lapply(ph_coefs, function(x) {
  two_vars <- all(dim(x) == c(100, 2))
  coef <- if (two_vars) {
    x[, 1]
  } else {
    x[1, ]
  }
  data.frame(
    coef = coef, rep = seq_along(coef)
  )
}
) |> dplyr::bind_rows(.id = "scenario")


coefs_res <- left_join(stata_coefs, icsp2_coef, by = c("scenario", "rep"))


library(ggplot2)
ggplot(
  coefs_res,
  aes(
    x = (b1 + coef)/2,
    y = b1 - coef,
  ) 
) + 
  geom_hline(colour = rgb(1, 0, 0, 0.5), yintercept = 0, linewidth = 20) +
  # annotate("ribbon", xmin = -Inf, x=0:1, ymin = -0.0001, ymax = 0.0001, fill = rgb(1, 0, 0, 0.5)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~scenario, nrow = 2) +
  theme_bw() +
  labs(
    x = "(stata + R)/2",
    y = "stata - R"
  )
ggsave("icsp2_v_stata.png", dpi = 600, width = 9.15, height = 4.32, units = "in")



ph_summary <- lapply(seq_along(scenarios), function(j) {
  coefs <- ph_coefs[[j]]
  sc <- scenarios[[j]]
  true_vals <- sc$b_trt[1]
  if (!is.null(sc$b_age)) true_vals <- c(true_vals, sc$b_age)
  data.frame(
    scenario = names(scenarios)[j],
    covariate = colnames(coefs),
    icsp2_mean = colMeans(coefs),
    icsp2_sd = apply(coefs, 2, sd),
    true_value = true_vals
  )
})
ph_summary <- do.call(rbind, ph_summary)
rownames(ph_summary) <- NULL

# Merge Stata results if available
if (!is.null(stata_coefs) && nrow(stata_coefs) > 0) {
  stata_summary <- lapply(split(stata_coefs, stata_coefs$scenario), function(d) {
    b_cols <- grep("^b", names(d), value = TRUE)
    data.frame(
      scenario = d$scenario[1],
      covariate = b_cols,
      stata_mean = colMeans(d[b_cols], na.rm = TRUE),
      stata_sd = apply(d[b_cols], 2, sd, na.rm = TRUE)
    )
  })
  stata_summary <- do.call(rbind, stata_summary)
  rownames(stata_summary) <- NULL
  ph_summary <- merge(ph_summary, stata_summary, by = c("scenario", "covariate"),
    all.x = TRUE
  )
}

ph_summary
```

### Baseline survival comparison plots

```{r baseline-comparison-plots, fig.width=8, fig.height=10}
par(mfrow = c(4, 2), mar = c(4, 4, 2, 1))

for (j in seq_along(scenarios)) {
  sc <- scenarios[[j]]
  sc_name <- names(scenarios)[j]
  fit <- ph_results[[j]][[1]]

  strata_pars <- sc$strata_params
  if (is.null(strata_pars)) strata_pars <- list(list(shape = 2, scale = 2))
  n_strata <- length(strata_pars)

  x_max <- max(unlist(lapply(fit$intervals, function(int) max(int[is.finite(int)]))))

  plot(fit,
    main = paste0("S", j, ": ", sc$label),
    xlim = c(0, x_max),
    col = rep(seq_len(n_strata), each = 2),
    lwd = 2
  )

  # True baseline
  t_seq <- seq(0, x_max, length.out = 200)
  for (s in seq_len(n_strata)) {
    true_surv <- pweibull(t_seq, shape = strata_pars[[s]]$shape,
      scale = strata_pars[[s]]$scale, lower.tail = FALSE)
    lines(t_seq, true_surv, col = s, lty = 2, lwd = 2)
  }

  # Overlay SAS NPMLE if available
  if (!is.null(sas_npmle)) {
    sas_sc <- sas_npmle[sas_npmle$scenario == sc_name, ]
    if (nrow(sas_sc) > 0) {
      lines(sas_sc$time, sas_sc$survival, col = "blue", lty = 3, lwd = 2)
    }
  }

  # Overlay Stata baseline if available
  if (!is.null(stata_baseline)) {
    stata_sc <- stata_baseline[stata_baseline$scenario == sc_name, ]
    if (nrow(stata_sc) > 0) {
      lines(stata_sc$time_grid, stata_sc$surv0, col = "red", lty = 4, lwd = 2)
    }
  }

  legend("topright",
    legend = c("icsp2", "True", "SAS", "Stata"),
    lty = c(1, 2, 3, 4), col = c("black", "black", "blue", "red"),
    lwd = 2, cex = 0.6, bty = "n"
  )
}
```

## IDU Dataset Comparison (Stata)

```{r idu-comparison}
library(haven)
idu <- read_dta("./local/idu.dta")

idu_fit <- ic_sp_ph(
  Surv(ltime, rtime, type = "interval2") ~ age_mean +
    male +
    needle +
    inject +
    jail,
  data = idu
)
idu_res <- summary(idu_fit)
within(as.data.frame(idu_res$coef_matrix), {
  HR <- exp(Estimate)
  HR_u <- exp(Estimate + 1.96 * Std.Error)
  HR_l <- exp(Estimate - 1.96 * Std.Error)
})
```
